---
title: "Willingness to have kid in relation to income, age and number of existing kids"
author: "Yuhan Huang, Tong Li, Chong Zhang, Chuyi Zhang"
date: "2020-10-19T21:49:57-07:00"
categories: ["R"]
tags: ["R Markdown", "plot", "regression"]
output:
  html_document: default
  pdf_document: default
---

## Abstract
People's reproducing decisions could be quite influential towards family 
structure and social development under a more general concept. Therefore, this 
research uses survey data from the 2017 Canadian General Social Survey (GSS) to 
answer the following questions. How do the income level, number of existing 
kids, and couples' average age influence their willingness to have kids in the 
future? More specifically, how do these variables strengthen or weaken couples' 
willingness to have a child in the future?After presenting several statistical 
models (such as the logistic model), the question was discussed and interpreted 
based on the model outcome.

## Introduction
The topic of this study is about whether a family will want a child in the
future(FUTURE_CHILD), considering INCOME, AGE, and TOTAL_CHILDREN as 
predictor variables. In the meantime, we can find the most influential factor 
in this decision. The insight for this study is based on the diminishing birth 
rate year by year, the birth rate for 2017 was only 10.55%, and it has been 
decreasing since 2009 right after the financial crisis in 2008, which is the 
reason we(this research) include family income as a predictor variable. GSS's
primary purpose is to
provide information on specific social policy issues. This data analysis will 
show the severity of this problem, which would help enact new policies toward 
this issue.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(blogdown)
data <- read_csv('2020-10-20-gss.csv')
```


## Data
The data in this research is the survey data from the 2017 GSS. Combining the
research question following data variables were selected for the first 
(logistic) model: respondent's age, family income, the total number of children 
born in the past, and their willingness to have a child in the future. By 
omitting each variable's unknown value, 4086 remaining observations became 
accessible for the model(logistic model). During the cleaning process, the 
income level was defined from the 'Income_family' variable instead of 
'income_respondent' because, in an existing family, total or average income can 
better reflect the financial status than using the estimator's income. The 
dependent variable of 'future_children_intention' was modified into a binary 
outcome of '0' as 'probably not' or 'definitely not' to the question and '1' 
representing 'probably yes' and 'definitely yes' perform better under a logistic
model. The data could have missed some lurking variables, such as the age of the
existing children. However, the marginal change in willingness dependent on this
variable can be ignored because the combination of age and number of the current
children can indirectly reflect the change in children's age.

```{r}
working_data <- data %>%
  select('total_children','children_in_household', 'future_children_intention'
         ,'age','income_family')
```


```{r, markup = FALSE}
working_data1 <- working_data %>%
  mutate(future_child = future_children_intention == 'Probably yes'
         | future_children_intention == 'Definitely yes' ) %>%
  filter(40>=age & age>=16) %>%
  select(-'future_children_intention') %>%
  na.omit(working_data1)


#give an income level based on variable income_family
working_data1 <- working_data1 %>% 
  mutate(income_level = income_family)
lv1 <- replace(working_data1$income_level, 
               working_data1$income_level == "Less than $25,000", 1)
working_data1 <- working_data1 %>% mutate(income_level = lv1)

lv2 <- replace(working_data1$income_level, 
               working_data1$income_level == "$25,000 to $49,999" , 2)
working_data1 <- working_data1 %>% mutate(income_level = lv2)

lv3 <- replace(working_data1$income_level, 
               working_data1$income_level == "$50,000 to $74,999" , 3)
working_data1 <- working_data1 %>% mutate(income_level = lv3)

lv4 <- replace(working_data1$income_level, 
               working_data1$income_level == "$75,000 to $99,999" , 4)
working_data1 <- working_data1 %>% mutate(income_level = lv4)

lv5 <- replace(working_data1$income_level, 
               working_data1$income_level == "$100,000 to $ 124,999" , 5)
working_data1 <- working_data1 %>% mutate(income_level = lv5)

lv6 <- replace(working_data1$income_level, 
               working_data1$income_level == "$125,000 and more" , 6)
working_data1 <- working_data1 %>% mutate(income_level = lv6)
working_data1$income_level <- as.numeric(working_data1$income_level)
```


## Model

This research uses logistic regression to analyze the result, which is a binary 
dependent variable (yes or no). It is because the logistic regression model is 
used for predictive analysis to explain the relationship between one dependent 
binary variable (yes or no) and three nomial variables (age, income level, and 
total children).

```{r}
logmodel<-glm(future_child ~ age + total_children + income_level, 
              data = working_data1 , family = 'binomial')
summary(logmodel)
```


The regression equation of our study is: 
$$log(p/1-p)= 5.14 - 0.168x_age - 1.110x_total.children - 0.016x_income.level$$
For every additional unit increase in age we expect the log odds of 
willingness to have kid to decrease by 0.168.
For every additional unit increase in total_children we expect the log odds of 
willingness to have kid to have kid to decrease by 1.110.
For every additional unit increase in income_level we expect the log odds of 
willingness to have kid to decrease by 0.016.



```{r}
new_working_data <- 
  data.frame(age=seq(min(working_data1$age), 
                     max(working_data1$age), len = 4086), 
             total_children=seq(min(working_data1$total_children), 
                                max(working_data1$total_children), len = 4086),
                        
income_level = seq(min(working_data1$income_level),
                   max(working_data1$income_level), len = 4086))

new_working_data$future_child = 
  predict(logmodel, newdata=new_working_data, type="response")

par(mfrow=c(2, 2)) # put all plot together
plot(future_child ~ age, data = working_data1, col="red4")
lines(future_child ~ age, new_working_data, col="green4", lwd=2)

plot(future_child ~ total_children, data = working_data1, col="red4")
lines(future_child ~ total_children, new_working_data, col="green4", lwd=2)

plot(future_child ~ income_level, data = working_data1, col = "red4")
lines(future_child ~ income_level, new_working_data, col = "green4", lwd= 2)
```

## Results

The first plot examines the relationship between FUTURE_CHILD and AGE. The
equation is written as the probability of wanting a child in the future. The 
response value of 1 on the y-axis represents "yes," and 0 represents "no" 
(same for all four plots, omitted in the following paragraphs). The plot shows 
that the probability of wanting a child in the future decreases as the AGE 
increases. When the AGE is between 16 to 20, the line's slope is not very steep,
which indicates that the probability of wanting a child decreases slowly as AGE 
increases in this period. The line is more vertical for AGE between 20 to 25, 
which suggests a change in the AGE of 1 year has a more massive effect in this 
range. When the FUTURE_CHILD goes to zero at the AGE approaches 27, the line 
flattens.

The second plot examines the relationship between FUTURE_CHILD and 
TOTAL_CHILDREN. The plot shows that the probability of wanting a child in the 
future decreases as the TOTAL_CHILDREN increases. When the number of children 
in the family is between 0 and 1, the slope of the line is not very steep, which 
indicates that the probability of wanting a child decreases slowly among 0 to 1 
child. The line is steeper in the TOTAL_CHILDREN between 2 to 3, which indicates
that a change in TOTAL_CHILDREN increased by 1 has a larger effect in this 
range. When the FUTURE_CHILD goes to zero at the TOTAL_CHILDREN around 3, the 
line flattens.

Last plot examines the relationship between FUTURE_CHILD and INCOME_LEVEL. 
The plot shows that the probability of wanting a child in the future decreases 
as the INCOME_LEVEL increases. The line is steeper when INCOME_LEVEL is around 
less than $25,000 to $74,999, which indicates that the income range between 
$25,000 and $74,999 has a larger effect when making the decision of whether to 
have a child in the future. When the FUTURE_CHILD goes to zero at the 
INCOME_LEVEL beyond $75,000, the line flattens.


## Discussion
The GSS sample is drawn using an area probability design that randomly selects
respondents in households across Canada to participate in the survey. The 
general population would be all the Canadians. Respondents that become part of 
the GSS sample are from a mix of urban, suburban, and rural geographic areas, 
which are the frame population. The GSS uses Statistics Canada’s standard 
telephone frame, which combines landline and cellular telephone numbers from the
Address Register, the Census of Population, and various administrative sources, 
and has been integrated with Statistics Canada’s standard dwelling frame. Last, 
the sampled population is the people who actually participated in the survey, 
and the size is approximately 20,000 respondents by 2019. Since 2013, GSS has 
been collected using a combination of self-completed online questionnaires and 
telephone interviews. Participation in the study is strictly voluntary, and the 
sample for the GSS is based on the method of cluster sampling, so the standard 
errors are larger for the GSS than the simple random sampling. The non-sampling 
errors include inadequate frame-coverage error, non-response problem; response 
error; interviewer bias; and processing, editing, and entering error.

# Weaknesses
In this survey GSS, most of the data is categorical instead of numerical, which 
produces limitation of choices of models. The categorical data type produces 
disadvantages, however, considering topic like how a person makes a choice on 
the question, a binary outcome is able to explain the case better. 


# Next Steps
Working beyond this research, this data set of GSS can tell more about the 
decision-making question of having a child in the future or not. For example, 
under measuring the family's financial status, this research used 
'income_family'. From another perspective, 'income_respondent' can be combined 
with 'income_family' to show how answers differ between the person who bears 
more of the family's financial pressure and the rest of the members. A follow-up
survey could include questions like "Do you feel that you bear more financial 
pressure from the family or it is equally shared". It can tell something about 
the lurking variable of their decision to have a child. In the meantime, it also 
gives more possibility to provide a weighted stratified sampling based on their 
share of family financial pressure.

## References
1. https://stackoverflow.com/questions/36685921/plot-logistic-regression-curve-in-r/36686468

2. http://www.sthda.com/english/articles/32-r-graphics-essentials/126-combine-multiple-ggplots-in-one-graph/

3. cleaning data by Rohan Alexander and Sam Caetano

4. "Hello R Markdown" by Frida Gomam