---
title: "Multilevel prediction of 2020 American presidential election"
author: "Yuhan Huang, Tong Li, Chong Zhang, Chuyi Zhang"
date: "2020-11-02T00:00:00-07:00"
output:
  html_document:
    df_print: paged
  pdf_document: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE, results= "hide")
```
```{r}
library(tidyverse)
library(broom)
library(brms) 

library(lme4)

# Loading in the cleaned survey Data
survey_data <- read_csv("2020-11-02-survey_data.csv")

# Loading in the cleaned census Data
census_data <- read_csv("2020-11-02-census_data.csv")

census_data <-census_data %>%
  mutate(gender = toupper(sex))
survey_data$gender<-toupper(survey_data$gender)
```

# Multilevel prediction of 2020 American presidential election

## Yuhan Huang, Tong Li, Chong Zhang, Chuyi Zhang
## November 2nd, 2020

# Abstract

Due to Donald Trump's special political performance in the past four years, the result of the 2020 US presidential election has become significantly focused by not only Americans but also the whole world. This report has used data from Democracy Fund + UCLA Nationscape to build a multilevel logistic regression model and then weight the estimate for each group according to the data from IPULM USA to predict the final result of the election through the post-stratification method. 

## Introduction
We downloaded the full dataset from Democracy Fund + UCLA Nationscape, the largest public opinion survey ever conducted. It includes people in every congressional district and mid-sized U.S. city in the leadup to the 2020 election. As we are approaching the presidential election date, nearly everyone worldwide is wondering what will be the result for 2020. We are thinking about predicting the outcome(whether Trump or Biden will win the election) based on the data we have right now. Another goal is finding any voting patterns among these individuals based on their gender, age, education level, and specifically the census region.

# Data
The data of this research was collected by Voter Study Group and IPUMS USA.
The survey data from the Voter Study Group was cleaned by selecting related variables
and creating a new numeric variable recording if the survey respondent will vote for 
Donald Trump in the year 2020. The target population was all the Americans who are potential voters for the presidential election; the frame population was American citizens who had the right to vote, and the sampled population was the people who actually participated in the survey online. We used a survey study method to conduct the sample, and the post-stratification data was cleaned by counting the number of people in each cell to result in the weighted estimating process. 
 


## Model

# Model Specifics
Due to the interest in the 2020 American presidential election result. This research
chose the Multilevel logistic model to predict election result for each 
individual and uses the post-stratification technique to weight the estimates 
to return the final prediction of the election outcome since MRP is often used 
for dichotomous fixed choice questions. We use this model due to the goal of 
estimating the odds that whether Trump will win the election(the yes/no outcome)
while taking the dependency of data into account (the fact that voting intention
is not finalized). Practically, it allows us to estimate such odds as a function
of lower level variables (age, gender), higher level variables (education level,
census regions), and the way they are interrelated.
By fitting in all the variables, we get the 
$$ y_{trump}=\beta_{census.region}+0.547x_{male}+0.076x_{old}-0.43x_{young}-0.02x_{higher.education}$$
where $ \beta_{census.region}=-0.46327-0.105w_{north}+0.195w_{south}-0.143w_{west}$
We can tell from the beta that gender and age play an important role when 
deciding the voting, census_region, also relatively high contribution to the 
overall model.

```{r,results='markup'}
survey_data_new <- survey_data %>%
  select(gender,
         vote_2020,
         vote_2016,
         age,
         education,
         census_region)%>%
  mutate(vote_trump_2020= vote_2020=='Donald Trump')%>%
  mutate(vote_trump_2016= vote_2016=='Donald Trump')%>%
  mutate(vote_Biden_2020= vote_2020=='Joe Biden')%>%
  mutate(higher_education = education == 'College Degree (such as B.A., B.S.)'
         |education == 'Doctorate degree'|education == 'Associate Degree'
         |education == 'Masters degree'|education ==
           'Completed some college, but no degree'
         |education =='Completed some graduate, but no degree')%>%
  filter(age>=18)
survey_data_new$age_1[survey_data_new$age >=18 ]<-"Old_adults"
survey_data_new$age_1[survey_data_new$age <= 60]<-"Mid_age"
survey_data_new$age_1[survey_data_new$age <= 40]<-"Young_adults"
survey_data_new$age<-survey_data_new$age_1
model_logit <- glmer(vote_trump_2020 ~ gender+
         age+higher_education+(1|census_region),
                   data = survey_data_new, 
                   family = "binomial")
model_logit_b <- glmer(vote_Biden_2020 ~ gender+
         age+higher_education+(1|census_region),
                   data = survey_data_new, 
                   family = "binomial")
model_logit1 <- glm(vote_trump_2020 ~ vote_trump_2016,
                   data = survey_data_new, 
                   family = "binomial")
summary(model_logit)

```

# Post-Stratification
Poststratification is a popular method for correcting known differences between 
sample and target populations (Little, 1993).  We post-stratified the census 
data responses to mimic the survey sample of likely voters. To make inference 
about the population level, we use the result of our model and apply the 
probability of voting for Trump to the population level, where $$y_j$$ is the 
estimate of y in cell j, and $N_j$ is the size of each $j^{th}$ cell in the 
population. Through the combination of different ages group(young_adults, 
middle-aged adults, old adults), citizens(naturalized citizens, not citizens),
gender(male, female), an education level(higher education, no higher education), 
and their census region(Midwest, Northeast, South, West), thus partitioning the 
data into 96 cells, results in the weighted estimating process. We focus on 
these four explanatory variables due to its influence on the voting result. For
instance,we chose the census_region as level 2, because different states have 
policies that are more beneficial for specific groups, which will do Trump much
good effects with the electorate. We will transform the result of our model, 
for each cell we have $\hat{y}$ to calculate \[\hat{y^{ps}}=\frac{\sum{N_i}*\hat{y_j}}{\sum{N_i}}\]. 
This results in 41.7% supporting Trump, and following a similar process we get 
43.4% support for Biden. 

```{r}
census_data <- census_data %>% 
  mutate(education = educd)
census_data_new <- census_data
census_data_new <- census_data_new %>% 
  mutate(man = ifelse(gender == "MALE", 1, 0))%>%
  mutate(higher_education = education == "master's degree"
         |education == "associate's degree, type not specified"
         |education == "bachelor's degree"
         |education == "doctoral degree"
         |education =="professional degree beyond a bachelor's degree"
         |education =="some college, but less than 1 year"
         |education =="1 or more years of college credit, no degree")
census_data_new$census_region[census_data_new$region == 
                                "middle atlantic division"
                              |census_data_new$region == 
                                "new england division"]<-"Northeast" 
census_data_new$census_region[census_data_new$region == 
                                "east south central div"|census_data_new$region 
                              == "west south central div"|census_data_new$region
                              == "south atlantic division"]<-"South" 
census_data_new$census_region[census_data_new$region == 
                                "mountain division"|census_data_new$region == 
                                "pacific division"]<-"West"
census_data_new$census_region[census_data_new$region == 
                                "west north central div"|census_data_new$region
                              == "east north central div"]<-"Midwest"
census_data_new$age_1[census_data_new$age >=18 ]<-"Old_adults"
census_data_new$age_1[census_data_new$age <= 60]<-"Mid_age"
census_data_new$age_1[census_data_new$age <= 40]<-"Young_adults"
census_data_new$age<-census_data_new$age_1

```


```{r}
census_data_new$estimate <-
  model_logit %>%
  predict(newdata = census_data_new, type = "response")

result <- census_data_new%>%
  filter(age >= 18)%>%
  mutate(alp_predict_prop = estimate*n/sum(n)) %>%
  summarise(Vote_for_Trump = sum(alp_predict_prop))

print(result)


census_data_new$estimate <-
  model_logit_b %>%
  predict(newdata = census_data_new, type = "response")

result_b <- census_data_new%>%
  filter(age >= 18)%>%
  mutate(alp_predict_prop = estimate*n/sum(n)) %>%
  summarise(Vote_for_Biden = sum(alp_predict_prop))
print(result_b)
```
```{r,results='markup'}
data_table <- matrix(c("41.7%", "43.4%"), ncol = 2, byrow = TRUE)
colnames(data_table) <- c("Trump", "Biden")
rownames(data_table) <- c("Yhat ps")
data_table <- as.table(data_table)
data_table
```

## Results
According to the result of post-stratification Biden was predicted to have more support rate among all voters. This support rate advance may lead Biden to be elected. Of 6101 qualified participation in the survey,  about 2544 people will vote for Trump, 2648 will vote for Biden, and 909 citizens will not vote for either of them. We found out classifying people according to their age and census_region, both groups do not have an overall trend toward the model, due to the large p-value. But the classification of young adults does have a significant contribution toward the final result.


## Discussion
We finish the primary goal of estimating whether Trump will win the election. 
Besides that, we are also interested in voting patterns based on people’s
classification. The exciting thing is that people with higher education(earn 
bachelor or master’s degrees)tend to be against Trump; while individuals who 
have lower education levels tend to favor Trump. We choose age and sex as two 
strongly correlated explanatory variables to voting preferences, but we are not
sure about other variables. In the beginning, we thought census_region might 
make a massive difference between people’s choices, as it turned out, people’s 
voting behavior has little relation to their region.

# Weaknesses
Like all the survey estimates, these numbers are subject to a degree of error.
There is an even higher possibility in this case because we are asking 
respondents about future plans, which might change. As such, the estimates 
should be interpreted with these uncertainties in mind. Specifically, Donald 
Trump supporters(registered voters) intend to vote in-person on Election Day, 
which is a much higher uncertainty towards the final result. 
Since we only use four explanatory variables out of 16 variables we could use, 
and the sample size is relatively large, our model is probably underfitting, 
too generalized for such a big dataset.


#Next Steps
When the election is done, we will have the finalized data. First we can check 
whether we have the correct prediction by our model, we can check this by doing
a post-hoc analysis or another survey of how to better improve estimation for 
future elections. Further studies may focus on voting patterns across the 
election years. We can check whether people who voted for Trump in 2016 still 
voted for Trump in 2020. This can be further analyzed to gain more insight about
how satisfied Trump supporters are regarding the performance of the Trump
administration. 

## Reference
1. IPUMS USA, census data of 2018 USA, Retrieved from:
https://usa.ipums.org/usa-action/data_requests
2. Voter Study Group, Nationscape, Retrieved from:
https://www.voterstudygroup.org/downloads?key=4b0ccde1-6d71-4a6f-8c5b-69e8172e4688

